scRNAseq

### 2018-02-26 datasets/scrnaseq/scRNAseq-keegan.Rmd
Add checks for precomputed objects; Add scDD evaluation

---
title: "scRNAseq"
author: "Ayshwarya Subramanian"
date: "9/15/2017"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE,cache.lazy = FALSE)

```
## R workspace setup

@@

# useful libraries
library(R.utils)
library(knitr)
library(BiocParallel)

@@

## scRNAseq datasets

```{r echo=FALSE, results='asis'}
df=data.frame(datasets=c("GSE63818-GPL16791","GSE84465","GSE45719","GSE94383"),organism=c("human","human","mouse","mouse"),technology=c("smart-seq","smart-seq","smart-seq2","smart-seq2"))
kable(df,caption="scRNAseq Datasets")
df1=data.frame(DE_method=c("Wilcoxon-rank sum","ScDD","MAST"),input=c("tpm","count","tpm"),output=c("p-val","p-val","p-val"))
kable(df1,caption="DE methods")

We selected two datasets, one each from mouse and human spp, from the `Conquer` (Soneson et al., 2017) scRNAseq data collection. `Conquer` provides uniformly processed gene expression data summarized to both count (aggregated from transcript members), and length-scaled TPMs at the gene level. The authors use `salmon` to quantify raw reads and `tximport` to summarize to gene level. The data comes from Smartseq, a plate based sequencing method which results in deeper sequencing of reads at the cost of numbers of cells.

## Data Download

We first download the data in the data directory.`conquer` provides the processed datasets in convenient .rds files. First, let's download the Human datasets.

@@

Next, we download the Mouse data:GSE45719

@@

```{r, dataload}
# load MultiAssayExperiment
humandata <- readRDS(paste0(datdir,"GSE63818-GPL16791.rds"))

@@

# save SummarizedExperiment 
saveRDS(humandata, paste0(datdir, "h1_GSE63818-GPL16791.rds"))
```

@@

```{r}
# load MultiAssayExperiment
humandata2 <- readRDS(paste0(datdir, "GSE84465.rds"))

@@

# save SummarizedExperiment 
saveRDS(humandata2, paste0(datdir, "h2_GSE84465.rds"))
```

@@

```{r,mouse}
# load MultiAssayExperiment
mousedata=readRDS(paste0(datdir, "GSE45719.rds"))

@@

# save SummarizedExperiment 
saveRDS(mousedata, paste0(datdir, "m1_GSE45719.rds"))
```

@@

```{r}
# load MultiAssayExperiment
mousedata2=readRDS(paste0(datdir,"GSE94383.rds"))

@@

# save SummarizedExperiment 
saveRDS(mousedata2, paste0(datdir, "m2_GSE94383.rds"))
```

@@

```{r, wilcoxon}
# function to add wilcox p-values
compute_wilcoxde_sc <- function(dat){
  expr <- assays(dat)[["TPM"]] 
  tmm <- edgeR::calcNormFactors(expr)
  expr <- edgeR::cpm(expr, lib.size = tmm * colSums(expr))
  rowData(dat)$wilcox_p <- sapply(seq_len(nrow(expr)), 
    function(i) {
      wilcox.test(expr[i, ] ~ colData(dat)$group)$p.value
    })
  return(dat)
}

@@

mousedata2 <- compute_wilcoxde_sc(mousedata2)
```

### Differences in distributions of gene expression in the 2 conditions
** Still need to clean up from here on; nothing below is eval'd**
We apply scDD to log-transformed counts as below. Th helper functions set up the Single-Cell experiment object.

```{r, scdd, eval=FALSE}

## human data
g_cdata_sub=readRDS(paste0(datdir,"h_g_nofilter_subcdata_GSE63818-GPL16791.rds"))
compute_scdd_sc(g_cdata_sub, h_scdd_dat1)

g_cdata_sub=readRDS(paste0(datdir,"h_g_nofilter_subcdata_GSE84465.rds"))
compute_scdd_sc(g_cdata_sub, h_scdd_dat2)

## mouse data
g_cdata_sub=readRDS(paste0(datdir,"m_g_nofilter_subcdata_GSE45719.rds"))
compute_scdd_sc(g_cdata_sub, m_scdd_dat1)

g_cdata_sub=readRDS(paste0(datdir,"m_g_nofilter_subcdata_GSE94383.rds"))
compute_scdd_sc(g_cdata_sub, m_scdd_dat2)

@@

### MAST

>>>>

---
title: "scRNAseq"
author: "Ayshwarya Subramanian and Keegan Korthauer"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE,cache.lazy = FALSE)
```
# Summary

```{r echo=FALSE, results='asis'}
library(knitr)
df=data.frame(datasets=c("GSE63818-GPL16791","GSE84465","GSE45719","GSE94383"),
              organism=c("human","human","mouse","mouse"),
              technology=c("smart-seq","smart-seq","smart-seq2","smart-seq2"),
              comparison=c("Somatic vs Primordial Germ Cells", 
                           "Neoplastic vs Regular",
                           "16-cell stage blastomere vs Mid blastocyst cell",
                           "LPS vs No stimulation"))
kable(df,caption="scRNAseq Datasets")
df1=data.frame(DE_method=c("Wilcoxon-rank sum","ScDD","MAST"),
               input=c("tpm","count","tpm"),
               output=c("p-val","p-val","p-val"))
kable(df1,caption="DE methods")
```

We selected four datasets, two each from mouse and human. 
These datasets come from the `Conquer` (Soneson et al., 2017) scRNAseq data collection. `Conquer` provides uniformly processed gene expression data summarized to both count (aggregated from transcript members), and length-scaled TPMs at the gene level. The authors used `salmon` to quantify raw reads and `tximport` to summarize to gene level. The data comes from Smartseq, a plate based sequencing method which results in deeper sequencing of reads at the cost of numbers of cells.

Here we are interested in performing differential expression analyses to determine
which

# Workspace setup

@@

# useful libraries
library(R.utils)
library(BiocParallel)

@@

# Data preparation

We first download the data in the data directory.`conquer` provides the processed datasets in convenient .rds files. First, let's download the Human datasets: GSE63818-GPL16791 and
GSE84465.

@@

Next, we download the Mouse data: GSE45719 and GSE94383

@@

```{r, dataload}
datfile_h1 <- paste0(datdir, "h1_GSE63818-GPL16791.rds")

if (!file.exists(datfile_h1)){

   # load MultiAssayExperiment
  humandata <- readRDS(paste0(datdir,"GSE63818-GPL16791.rds"))

@@

  # save SummarizedExperiment 
  saveRDS(humandata, datfile_h1)
}else{
  humandata <- readRDS(datfile_h1)
}
```

@@

```{r}
datfile_h2 <- paste0(datdir, "h2_GSE84465.rds")

if (!file.exists(datfile_h2)){

  # load MultiAssayExperiment
  humandata2 <- readRDS(paste0(datdir, "GSE84465.rds"))

@@

  # save SummarizedExperiment 
  saveRDS(humandata2, datfile_h2)
}else{
  humandata2 <- readRDS(datfile_h2)
}
```

@@

```{r,mouse}
datfile_m1 <- paste0(datdir, "m1_GSE45719.rds")

if (!file.exists(datfile_m1)){

  # load MultiAssayExperiment
  mousedata=readRDS(paste0(datdir, "GSE45719.rds"))

@@

  # save SummarizedExperiment 
  saveRDS(mousedata, datfile_m1)
}else{
  mousedata <- readRDS(datfile_m1)
}
```

@@

```{r}
datfile_m2 <- paste0(datdir, "m2_GSE94383.rds")

if (!file.exists(datfile_m2)){

  # load MultiAssayExperiment
  mousedata2=readRDS(paste0(datdir,"GSE94383.rds"))

@@

  # save SummarizedExperiment 
  saveRDS(mousedata2, datfile_m2)
}else{
  mousedata2 <- readRDS(datfile_m2) 
}
```

@@

```{r, wilcoxon}
# function to add wilcox p-values
compute_wilcoxde_sc <- function(dat){
  # check if already computed
  if (! "wilcox_p" %in% colnames(rowData(dat))){
    expr <- assays(dat)[["TPM"]] 
    tmm <- edgeR::calcNormFactors(expr)
    expr <- edgeR::cpm(expr, lib.size = tmm * colSums(expr))
    rowData(dat)$wilcox_p <- sapply(seq_len(nrow(expr)), 
      function(i) {
        wilcox.test(expr[i, ] ~ colData(dat)$group)$p.value
      })
  }else{
     message("Wilcox p-values already computed.")
  }
  return(dat)
}

@@

mousedata2 <- compute_wilcoxde_sc(mousedata2)

# save results
saveRDS(humandata, datfile_h1)
saveRDS(humandata2, datfile_h2)
saveRDS(mousedata, datfile_m1)
saveRDS(mousedata2, datfile_m2)
```

### Differences in distributions of gene expression in the 2 conditions

We apply scDD to normalized log-transformed counts.

```{r, scdd, eval=FALSE}
# function to add scDD p-values
compute_scdd_sc <- function(dat){
  # check if already computed
  if (! "scdd_p" %in% colnames(rowData(dat))){
    sce <- SingleCellExperiment(assays=list(count=assays(dat)[["count"]]),
                                colData=data.frame(colData(dat)))
    # scran normalization
    libsize <- scran::computeSumFactors(assay(sce, "count"))
    normcounts(sce) <- GetNormalizedMat(assay(sce, "count"), libsize)

    rowData(dat)$scdd_p <- 
        results(scDD(sce, 
             condition = "group",
             prior_param = list(alpha=0.01, mu0=0, s0=0.01, a0=0.01, b0=0.01), 
             testZeroes = FALSE))$nonzero.pvalue
  }else{
     message("scDD p-values already computed.")
  }
  return(dat)
}

## human data

compute_scdd_sc(humandata)

compute_scdd_sc(humandata2)

compute_scdd_sc(mousedata)

compute_scdd_sc(mousedata2)

@@

### MAST
** Still need to clean up from here on; nothing below is eval'd**
